Equilivest: A Robotic Vest to aid in Post-Stroke Dynamic Balance Rehabilitation

### 2022-09-19 iros.tex
Almost ready.

\author{%
Franco Paviotti$^{1}$, Esteban Buniak$^{2}$, Rodrigo Ramele$^{3}$, Orestes Freixes$^{4}$ and Juan Miguel Santos$^{5}$% <-this % stops a space
\thanks{$^{1}$F. Paviotti is with Bioengineering, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small fpaviotti@itba.edu.ar}}%
\thanks{$^{2}$E.Buniak is with Ingeniería en Informática, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small ebuniak@itba.edu.ar}}%
\thanks{$^{3}$R.Ramele is with Ingeniería en Informática, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small rramele@itba.edu.ar}}%
\thanks{$^{4}$O. Freixes is with the CINER Centro Integral de Neurorehabilitación,
                    Buenos Aires, Argentina
        {\tt\small orestesfreixes@gmail.com}}%
\thanks{$^{5}$J.M. Santos is with the UNAHUR Universidad Nacional de Hurlingham,
                    Buenos Aires, Argentina
        {\tt\small juanmiguelsantos@gmail.com}}%
%\thanks{*This work was not supported by any organization}% <-this % stops a space}
}

@@

Patient is a 31 years old female, without any history of chronic ailment, who suffered an acute brainstem stroke after giving birth.  The stroke was on posterior fossa subarachnoid due to a brain artereovenous malformation (AVM), which could be linked to pregnancy or puerperium~\cite{Porras.etal2017}.  Patient was in coma for around 2 months, and after that unable to walk, move, talk or swallow.  After two years of intensive rehabilitation, Patient managed to recover significantly, including from dysphagia, which was very important in order to remove the feeding tube allowing her to start speech recovering.

After 24 months since event, the patient, was discharged from hospital and maintained a 3 times per week rehabilitation treatment, focusing on a remaining affection related to dynamic balance problems while walking.  The patient achieved satisfactory index scores in static balance tests and is able to perform static hip-balance and ankle-balance.  She has recovered muscle in her legs and can perform lower-limb exercises.  Her vision is normal.

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}[!t]{|p{0.45\linewidth} |p{0.45\linewidth} |}
\hline
\textbf{Question} & \textbf{Answer} \\
\hline
\hline
Under which situations do you feel you are prone to fall ? & When I am tired, distracted or stressed. \\
\hline
What do you feel ?  & I feel that I loss my balance\\
\hline
What do you feel before you fall ? & Sometimes I do feel something before.  I realize that I am going to fall, but there isn't anything I can do to avoid it. \\
\hline
When do you use the cane ?  & Only when I am out of home.\\
\hline
\end{tabular}
\vspace{2pt}
\caption{Survey and responses provided by the patient and family.}
\label{tab:patientsurvey}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\begin{tabular}[!t]{|p{0.45\linewidth} |p{0.45\linewidth} |}
\hline
\textbf{Question} &  \textbf{Answer} \\
\hline
\hline
Describe the situation when the patient loss her balance & Loss of balance is manifested at the beginning, during the gait cycle or with sudden stops. Walking is performed using intense visual compensation.   Instability increase when the patient is distracted, when she stops looking at the floor, or when she looks sideways.  Her gait is not automated, and demands cognitive resources. \\
\hline
Is there any significant muscle loss in lower-limbs ? & No\\
\hline
Does she presents any of the following dysmetria / heavy shaking / somatosensorial alterations / proprioceptive alterations ? & No \\
\hline
Does she have any visible reaction before falling?  & She presents dynamic alterations in hip and ankle compensation.\\
\hline
Is the patient gait normal ?  &  No, the patient presents ataxic gait, likely triggered by slow reaction to perform lower-limb balance correction. \\
\hline
Is there any behavioral pattern linked to falling events ? &  Yes, cognitive workload.  Dual tasks situations, when the patient needs to perform something extra while walking. \\
\hline
Describe current treatment. & The patient is currently working on rehabilitation exercises to retrain her Vestibulo-Ocular Reflex (VOR), to improve her static and dynamic balance, and improve her locomotor automatism.  She is improving on a monthly basis, but still she has not reached the level to walk autonomy without any aid. \\
\hline
\end{tabular}
\vspace{2pt}
\caption{Survey and responses provided by his main patient's rehabilitation caregiver.}
\label{tab:caregiverssurvey}
\end{center}
\end{table*}

We perform a series of surveys and interviews with the patient and their caregivers.  Main results are summarized in 
Tables ~\ref{tab:patientsurvey} and \ref{tab:caregiverssurvey}.  Based on the clinical history, results from surveys and interviews, we postulate three different potential clinical situations that could use the external biofeedback signal and could potentially aid in rehabilitation procedures.  The first (i) is a potential problem in the integration of vestibular information while walking, the second (ii) is bradykinesia where the required processing speed to effectively adjust the lower limb to keep the center of mass inside the base support is not achieved.  Finally, the third (iii) hypothesis is an ataxic gait, where non-automated gait produces an increase in the likelihood of failing.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{equilivest.png}
\caption{Front and back view of the Equilivest prototype.  The vest is based on a safety vest, with additional velcro pockets.  The battery is located on the back, while the vibrotactile motor is located on the belly~\cite{Brandebusemeyer.etal2021}.  The IMU sensor is pressed against the chest by the use of a flexible elastic band. }
\label{fig:smart-vest}
\end{figure}

@@

Preliminary results show that for the (1) experiment (Figure \ref{fig:vestibular}), the pitch,roll, yaw angles can be used to determine a breakpoint where the fall is inevitable (black vertical line, marked as 1).   Results from experiment number (2) in Figure \ref{fig:gait} show that the gait cycle are clearly visible from raw acceleration data.

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{vestibular.png}
\caption{Pitch value in degrees showing the progressive increase as participants leaned forward crossing the instability condition (1).  Time is in milliseconds. }
\label{fig:vestibular}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{gait.png}
\caption{Angular acceleration for the X and Y axis showing 7 steps of the 10MWT assessment.  Time is in milliseconds. }
\label{fig:gait}
\end{figure}

@@

The information provided from all the experiments allow us to implement three different VF strategies:  The first is an \textbf{Artificial Vestibular Feedback}, which provides information as a feedback for vestibular information integration process.  This is implemented as a frequency-increasing VF signal that represents a distance from the breaking condition position where the fall is inevitable.   The purpose is to provide the patient with a continuous sensation that modulates the risk of falling.

The second strategy is the implementation of \textbf{Gait Pacemaker}.  The obtained data from the analysis of the ataxic gait is used to implement a podometer.  It has been showed that gait synchronization with music increase gait performance~\cite{Roerdink.etal2007} and synchronized gait trainer achieved positive results on patient~\cite{Blicher.etal2009}.  Hence, the aim of this stimulation approach is to keep the patient as close as possible to a normal and safe gait.

@@

\section*{Conclusion}
Overall, this devices allows to implement a testbed that can be iteratively extended to obtain experimental data and to implement biofeedback strategies which could potentially lead to an increase in the effectiveness of different rehabilitation procedures~\cite{Bowman2021}.

>>>>

\author{%
Franco Paviotti$^{1}$, Esteban Buniak$^{2}$, Rodrigo Ramele$^{3}$, Orestes Freixes$^{4}$ and Juan Miguel Santos$^{5}$% <-this % stops a space
\thanks{$^{1}$F. Paviotti is with Bioengineering, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small fpaviotti@itba.edu.ar}}%
\thanks{$^{2}$E.Buniak is with Ingeniería en Informática, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small ebuniak@itba.edu.ar}}%
\thanks{$^{3}$R.Ramele is with Ingeniería en Informática, Instituto Tecnológico de Buenos Aires
                    Buenos Aires, Argentina
        {\tt\small rramele@itba.edu.ar}}%
\thanks{$^{4}$O. Freixes is with the CINER Centro Integral de Neurorehabilitación,
                    Buenos Aires, Argentina
        {\tt\small orestesfreixes@gmail.com}}%
\thanks{$^{5}$J.M. Santos is with the UNAHUR Universidad Nacional de Hurlingham,
                    Buenos Aires, Argentina
        {\tt\small juan.santos@unahur.edu.ar}}%
%\thanks{*This work was not supported by any organization}% <-this % stops a space}
}

@@

Patient is a 31 years old female, without any history of chronic ailment, who suffered an acute brainstem stroke after giving birth.  The stroke was on posterior fossa subarachnoid due to a brain artereovenous malformation (AVM), which could be linked to pregnancy or puerperium~\cite{Porras.etal2017}.  Patient was in coma for around 2 months, and after that unable to walk, move, talk or swallow.  After two years of intensive rehabilitation, the patient managed to recover significantly, including from dysphagia, which was very important in order to remove the feeding tube allowing her to start speech recovering.

After 24 months since event, the patient, was discharged from hospital and maintained a 3 times per week rehabilitation treatment, focusing on a remaining affection related to dynamic balance problems while walking.  The patient achieved satisfactory index scores in static balance tests and is able to perform static hip-balance and ankle-balance.  She has recovered muscle force in her legs and can perform lower-limb exercises.  Her vision is normal.

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}[!t]{|p{0.45\linewidth} |p{0.45\linewidth} |}
\hline
\textbf{Question} & \textbf{Answer} \\
\hline
\hline
Under which situations do you feel you are prone to fall ? & When I am tired, distracted or stressed. \\
\hline
What do you feel ?  & I feel that I loss my balance\\
\hline
What do you feel before you fall ? & Sometimes I do feel something before.  I realize that I am going to fall, but there isn't anything I can do to avoid it. \\
\hline
When do you use the cane ?  & Only when I am out of home.\\
\hline
\end{tabular}
\vspace{2pt}
\caption{Survey and responses provided by the patient.}
\label{tab:patientsurvey}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\begin{tabular}[!t]{|p{0.45\linewidth} |p{0.45\linewidth} |}
\hline
\textbf{Question} &  \textbf{Answer} \\
\hline
\hline
Describe the situation when the patient loss her balance & Loss of balance is manifested at the beginning, during the gait cycle or with sudden stops. Walking is performed using intense visual compensation.   Instability increase when the patient is distracted, when she stops looking at the floor, or when she looks sideways.  Her gait is not automated, and demands cognitive resources. \\
\hline
Does she presents any of the following muscular dystrophy in lower-limbs / dysmetria / heavy shaking / somatosensorial alterations / proprioceptive alterations ? & No \\
\hline
Does she have any visible reaction before falling?  & She presents dynamic alterations in hip and ankle compensation.\\
\hline
Is the patient gait normal ?  &  No, the patient presents ataxic gait, likely triggered by slow reaction to perform lower-limb balance correction. \\
\hline
Is there any behavioral pattern linked to falling events ? &  Yes, cognitive workload.  Dual tasks situations, when the patient needs to perform something extra while walking. \\
\hline
Describe current treatment. & The patient is currently working on rehabilitation exercises to retrain her Vestibulo-Ocular Reflex (VOR), to improve her static and dynamic balance, and improve her locomotor automatism.  She is improving on a monthly basis, but still she has not reached the level to walk autonomy without any aid. \\
\hline
\end{tabular}
\vspace{2pt}
\caption{Survey and responses provided by his main patient's rehabilitation caregiver.}
\label{tab:caregiverssurvey}
\end{center}
\end{table*}

We perform a series of surveys and interviews with the patient and their caregivers.  Main results are summarized in 
Tables ~\ref{tab:patientsurvey} and \ref{tab:caregiverssurvey}.  Based on these results and patient's clinical history, we postulate three different potential clinical situations that could use the external biofeedback signal eventually aiding in rehabilitation procedures.  The first (i) is a potential problem in the integration of vestibular information while walking, the second (ii) is bradykinesia where the required processing speed to effectively adjust the lower limb to keep the center of mass inside the base support is not achieved.  Finally, the third (iii) hypothesis is an ataxic gait, where non-automated gait produces an increase in the likelihood of failing.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{equilivest.png}
\caption{Front and back views of the Equilivest prototype.  The vest is based on a safety vest, with additional velcro pockets.  The battery is located on the back, while the vibrotactile motor is located on the belly~\cite{Brandebusemeyer.etal2021}.  The IMU sensor is pressed against the chest by the use of a flexible elastic band. }
\label{fig:smart-vest}
\end{figure}

@@

Preliminary results show that for the (1) experiment, on Figure \ref{fig:vestibular}), the value of pitch angles can be used to determine a breakpoint where the fall is inevitable (black vertical line, marked as 1*).   Results from experiment number (2) in Figure \ref{fig:gait} show that the gait cycle is clearly visible from raw acceleration data.

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{vestibular.png}
\caption{Pitch value in degrees and angular acceleration around X axis showing the progressive increase as participants leaned forward crossing the instability condition (1).  Time is in milliseconds. }
\label{fig:vestibular}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=8cm]{gait.png}
\caption{Angular acceleration for the X and Z axis showing 7 steps of the 10MWT assessment.  Time is in milliseconds. }
\label{fig:gait}
\end{figure}

@@

The information provided from all the experiments allow us to implement three different VF strategies:  The first is an \textbf{Artificial Vestibular Feedback}, which provides information as a feedback for vestibular information integration process.  This is implemented as a frequency-increasing VF signal that increases as long as the distance from the breaking condition position where the fall is inevitable decreases.   The purpose is to provide the patient with a continuous sensation that modulates the risk of falling.

The second strategy, independent of the first one, is the implementation of \textbf{Gait Pacemaker}.  The obtained data from the analysis of the ataxic gait is used to implement a podometer.  It has been showed that gait synchronization with music increase gait performance~\cite{Roerdink.etal2007} and synchronized gait trainer achieved positive results on patient~\cite{Blicher.etal2009}.  Hence, the aim of this stimulation approach is to keep the patient as close as possible to a normal and safe gait.

@@

\section*{Conclusion}
Overall, we aim for \textit{Equilivest} to serve as a testbed that can be iteratively extended to obtain experimental data and to implement biofeedback strategies~\cite{Bowman2021}.
